---
title: "In-Class: Lines & Polygons"
output: html_notebook
---

For this activity, we will focus on loading and manipulating `LINESTRING` and `POLYGON` objects derived from the City of Richmond, GIS Department.

Here are the URL's for the shapefile zip archives.

```{r}
roads_url <- "https://github.com/dyerlab/ENVS-Lectures/raw/master/data/Centerlines-shp.zip"
district_url <- "https://github.com/dyerlab/ENVS-Lectures/raw/master/data/Zoning_Districts-shp.zip"
schools_url <- "https://github.com/dyerlab/ENVS-Lectures/raw/master/data/rps_schools.zip"
```

And here is the CSV file for describing Zoning Types.

```{r}
zone_url <- "https://raw.githubusercontent.com/dyerlab/ENVS-Lectures/master/data/DistrictCodes.csv"
```

## Questions

For these questions, you may want to 

1. In which Zoning District is Monroe Park located and what is the area of that district.

2. Create a map of the area enclosed by N Belvidere St, East Broad St, and I-95.  Within this area, show the streets, but not alleys.

3. What kind of Zoning Districts are within this area?  Make a table showing category & area.

4. Download the Richmond School shapefile above and make a table of all the schools that are within 1000 ft of any highways in Richmond, Virginia.

```{r}

```




