---
title: "In-Class: Lines & Polygons"
output: html_notebook
---

```{r startup, include=FALSE}
library( sf )
library( units )
library( knitr )
library( kableExtra )
library( tidyverse )
theme_set( theme_minimal( base_size = 14))
knitr::opts_chunk$set( message = FALSE,
                       warning = FALSE,
                       echo = TRUE )
```



For this activity, we will focus on loading and manipulating `LINESTRING` and `POLYGON` objects derived from the City of Richmond, GIS Department.

Here are the URL's for the shapefile zip archives.  Let's first download them and then expand them in place (we will delete them at the end of the homework).

```{r}
roads_url <- "https://github.com/dyerlab/ENVS-Lectures/raw/master/data/Centerlines-shp.zip"
download.file( roads_url, destfile = "Centerlines-shp.zip" )
unzip( "Centerlines-shp.zip" )
st_read( "Centerlines-shp/tran_Carriageway.shp") %>% 
  select( FID,                                         # Select usable columns.
          `Street Name` = FullName, 
          `One Way` = OneWay, 
          Type = StreetType, 
          `Speed Limit` = PostedSpee,
          geometry) %>%
  mutate( `One Way` = ifelse( is.na(`One Way`), TRUE, FALSE), # Fix One Way & Type
          Type = factor( Type) ) %>%                     
  arrange( `Street Name`) -> rva_roads 


district_url <- "https://github.com/dyerlab/ENVS-Lectures/raw/master/data/Zoning_Districts-shp.zip"
download.file( district_url, destfile = "Zoning_Districts-shp.zip" )
unzip( "Zoning_Districts-shp.zip" )
unlink( "Zoning_Districts-shp.zip" )
st_read("Zoning_Districts.shp") %>% 
  select( OBJECTID,
          `Zoning Code` = Name, 
          geometry) -> rva_districts


schools_url <- "https://github.com/dyerlab/ENVS-Lectures/raw/master/data/rps_schools.zip"
download.file( schools_url, destfile = "rps_schools.zip" )
unzip( "rps_schools.zip" )
st_read( "f4996b83-ce9c-4eef-94d0-dc53e4e269e32020314-1-5d1dz8.ribms.shp" ) %>% 
  select( OBJECTID, 
          `School Name` = Name,
          Type, 
          Address,
          geometry) %>% 
  mutate( Type = factor( Type,       # set school type as factor
                         ordered=TRUE,
                         levels = c( "Specialty", "Elementary", "Middle", "High") ) ) -> rva_schools 
```

And here is the CSV file for describing Zoning Types.

```{r}
zone_url <- "https://raw.githubusercontent.com/dyerlab/ENVS-Lectures/master/data/DistrictCodes.csv"
read_csv( zone_url ) %>%
  mutate( `Zoning Code` = Name ) %>%
  select( -Name ) -> types 
rva_districts %>%
  left_join( types ) %>% summary() 
```

## Questions

For these questions, you may want to 

1. In which Zoning District is Monroe Park located and what is the area of that district.

*OK, so what I did is find the roads that encircle Monroe Park.  Then plot them to look at what is going on.*
```{r}
rva_roads %>% 
  filter( `Street Name` %in% c("N Laurel St",
                               "W Franklin St",
                               "W Main St",
                               "N Belvidere St")) -> monroe
plot( st_geometry( monroe ), axes=TRUE )
```

*OK, so let's zoom in to the part that the park is within just using the `xlim` and `ylim`.  I literally just copied off the axes of the figure above.*

```{r}
plot( st_geometry( monroe), 
      axes=TRUE,
      xlim=c(11784000, 11790000), 
      ylim=c(3723000, 3725000))
```

*I can now crop both roads and districts to this extent*

```{r}
monroe %>% 
  st_crop( xmin = 11784000, 
           xmax = 11790000,
           ymin = 3723000,
           ymax = 3725000) -> monroe_park
rva_districts %>%
  st_crop( monroe_park ) -> monroe_districts 
```


*Then I can plot the districts over the same place and label the objects by `OBJECTID` to identify which one encloses Monroe Park itself.*

```{r}
ggplot() + 
  geom_sf( aes( fill = `Zoning Code` ),
           data = monroe_districts ) + 
  geom_sf( color = "red",
           size = 1.25,
           data = monroe_park ) +
  geom_sf_text( aes( label = OBJECTID ),
                data = monroe_districts )
```

*So it is `OBJECTID==533`, which has an area of `r format( rva_districts %>% filter( OBJECTID==533 ) %>% st_area( m^2) %>% as.numeric(), digits=2, big.mark=",")` m<sup>2</sup>.*





2. Create a map of the area enclosed by N Belvidere St, East Broad St, and I-95.  Within this area, show the streets, but not alleys.

```{r}
rva_roads %>%
  filter( `Street Name` %in% c("I 95",
                               "N Belvidere St",
                               "E Broad St")) -> downtown

plot( st_geometry(downtown ) )
```
*OK, so it looks like there is a gap in `I-95` where it overlaps with "I-64".  Also, Broad Street goes from East to West in the target area.*

```{r}
rva_roads %>%
  filter( Type == "Highway" ) %>%
  st_drop_geometry() %>%
  select( `Street Name` ) %>%
  unique()
```

*OK, so let's grab the rest of the `I 95` streets by name and then graph it with axes on to be able to adjust the x- and y- axis scales to zoom in on that area of interest.  To do this, I litterally look at the axes and then iteratively plot and zoom in to get the area of interest in the forefront.*

```{r}
rva_roads %>%
  filter( `Street Name` %in% c("I 95",
                               "N Belvidere St",
                               "E Broad St",
                               "W Broad St",
                               "I 95 / I 64",
                               "I 95 N to Pocahontas Pkwy")) -> downtown

plot( st_geometry(downtown ), axes=TRUE, 
      xlim = c(11785500,11795000),
      ylim = c(3720000, 3728000))
```
*So now, we know what the area should be, we can plot all the data (minus the alleys).*

```{r}
st_crop( rva_roads, 
         xmin = 11786400, 
         xmax = 11795100,
         ymin = 3720400,
         ymax = 3726600 ) -> downtown_roads

st_crop( downtown, downtown_roads) -> downtown 

downtown_roads %>%
  filter( Type != "Alley" ) %>%
  ggplot() + 
  geom_sf( aes(color=Type)) + 
  geom_sf( data = downtown ) + 
  coord_sf( expand=FALSE,
            datum = st_crs(rva_roads) ) + 
  xlab("Easting") +
  ylab("Northing")
  
```


3. What kind of Zoning Districts are within this area?  Make a table showing category & area.

*Look to see which ones are in the box*

```{r}
ggplot() +
  geom_sf( data = rva_districts ) + 
  geom_sf( data = downtown,
           color = "red") + 
  geom_sf_text( aes(label=OBJECTID),
                data = rva_districts) + 
  coord_sf( xlim = c(11786500, 11795000),
            ylim = c(3720500, 3726500),
            expand=FALSE,
            datum = st_crs(rva_roads) ) + 
  xlab("Easting") +
  ylab("Northing")
```

*Now grab them and put them into a table.*


```{r}
rva_districts %>%
  filter( OBJECTID %in% c( 188, 187, 302,
                           11, 12, 216,
                           566, 565, 304,
                           303, 215, 564,
                           562, 314, 591) ) %>%
  mutate( Area = units::set_units(st_area(geometry), m^2 ) ) %>% 
  arrange( `Zoning Code` ) %>%
  st_drop_geometry() %>%
  select( `Zoning Code`, 
          `Object ID` = OBJECTID, 
          Area ) %>%
  kable( caption = "Table 1: Zoning Districts in downtown Richmond Virginia enclsed by Broad Street, Bevidere, and I-95." ) %>%
  kableExtra::kable_material()
```




4. Download the Richmond School shapefile above and make a table of all the schools that are within 1000 ft of any highways in Richmond, Virginia.


*So let's take a look a the raw data.  I'm going to plot the highways and highway interchanges with a 1000ft buffer (remember, the default datum for these data are ftUS).  Then overlay the schools on top of it.*

```{r}
rva_roads %>%
  filter( Type %in% c("Highway",
                      "Highway Interchange") ) %>%
  st_geometry() %>% 
  st_simplify() -> highways

plot( st_buffer( st_geometry(highways), dist=1000), col="grey90" )
rva_schools %>%
  select( Type, geometry ) %>%
  plot( add=TRUE, pch=16 )
```

*Looks like there are some schools here.  So let's do a spatial join using st_intersection and then plot it out as a table.*

```{r}
highways %>%
  st_buffer( dist = 1000 ) -> highway_buffer  # make buffers


# Filter schools
rva_schools %>%
  st_intersection( highway_buffer ) %>%
  arrange( Type, `School Name`) %>%
  select( `School Name`, Type, Address ) %>%
  st_drop_geometry() %>%
  knitr::kable(caption = "Table 2: Public schools that are within 1000 feet of a highway.") %>%
  kableExtra::kable_material()

```




### Cleaning up files.

We can clean up all the raw data hanging around by executing the following code.

```{r}
unlink("Zoning_Districts.*")
unlink("f4996b83-ce9c-4eef-94d0-dc53e4e269e32020314-1-5d1dz8.ribms.*")
unlink("Centerlines-sh*", recursive = TRUE, force=TRUE, expand = TRUE )
```










