---
title: "Point, Line, & Polygon In-Class"
output: html_notebook
---

This activity will focus on what we've learned thus far on how to work with `points`, `lines` and `ploygons` (and shapefiles) in `R`.

```{r}
library( sf )
library( tidyverse )
```

## The Data

The data for this consists of both zoning districts and road center lines for Richmond, Virginia.  These are available at:

```{r}
roads_url <- "https://github.com/dyerlab/ENVS-Lectures/raw/master/data/Centerlines-shp.zip"
district_url <- "https://github.com/dyerlab/ENVS-Lectures/raw/master/data/Zoning_Districts-shp.zip"
```


## The Big Picture

The main question you are going to work on here is:

> What roads intersect the zoning district that defines the heart of `The Fan` (hint, it is the one with OBJECTID=368)?  

For this in-class activity, you will also need to get a copy of the [Spatial Cheatsheet](https://github.com/rstudio/cheatsheets/raw/main/sf.pdf) to consult.



## The Process

The following questions will lead you through the process.

1. In list format, outline how you are going to answer the canonical question.  Be pedantic.

- Load in zip files
- ...

2. Load and process the zip archives so you can read in the two underlying shapefiles.

```{r}
# load and process here
```


3. Show that they are both in the same projection.


4. Identify and isolate the Zoning District of interest in `The Fan`.  

```{r}

```


5. Make a plot of the target zone as well as some around it (I'd add a small buffer to the bounding box).

```{r}

```


6. Looking at the [Spatial Cheatsheet](https://github.com/rstudio/cheatsheets/raw/main/sf.pdf), which function do you think would be most appropriate for finding the roads that cross OBJECTID=368?  Use `tidyverse` to identify the roads that cross this polygon and save it as a new `sf data.frame`

```{r}

```


7. Make a table of the unique road names in alphabetical order.

```{r}
library( knitr )
library( kableExtra )


```


8. Create a plot of the districts as the background, the target disctrict (OBJECTID=368) as highlighted in another color, and the roads that cross the it overlay on top.

```{r}

```




