---
title: "**Quantiative Lecture Content**"
author: "RJ Dyer"
date: "Center for Environmental Studies"
output:
  html_document:
    css: ["https://www.w3schools.com/w3css/4/w3.css", "./css/index_style.css"]
    toc: true
    theme: flatly
    toc_float: true
---


<!-- 
https://cs193p.sites.stanford.edu
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(purrr)
library(stringr)
library(tidyverse)
```

Below are all the lecture topics that I've developed for various classes taught at the [Center for Environmental Studies](https://ces.vcu.edu).  For more information on these lectures, feel free to contact me directly at [rjdyer@vcu.edu](mailto://rjdyer@vcu.edu).


```{r}
# Define some helper functions
btn <- function(path, color, name, as.nb=FALSE) {
  nb <- ifelse( as.nb, ".nb", "")
  code <- paste( "<a href='", 
                 path, "/",name, nb,
                 ".html' class='w3-button w3-round-small w3-", 
                 color, "'>", Hmisc::capitalize(name),"</a>",
                 sep="")
  return( code )
}

video <- function( path, color, name ) {
  val <- read_file(  paste(path,"video.txt",sep="/") )
  btn <- paste("<a href='",
               val,
               "' class='w3-button w3-round-small w3-", 
                 color, "'>",Hmisc::capitalize(name),"</a>", sep="")
  return(btn)
}

nonunderscorification <- function( files ) {
  lastPart <- matrix(unlist(str_split( files, "/")),ncol=2, byrow=TRUE)[,2]
  return(unlist( map( str_replace( str_replace_all(lastPart,"_", " "), "./", "" ),
                       gtools::capwords ) ))
}



makeTable <- function( files, caption = "" ) {
  buttons <- numeric(0)
  for( path in files ) {
    
    vid <- ifelse( file.exists(  paste(path,"video.txt", sep="/") ),
                   video( path, "amber", "video"), "")
    
    slides <- ifelse( file.exists(  paste(path,"slides.Rmd", sep="/") ),
                      btn(path,"orange","slides"),
                      "")
    
    narriative <- ifelse( file.exists(  paste(path,"narrative.Rmd", sep="/") ),
                          btn(path,"blue","narrative",as.nb = TRUE),
                          "")

    ##  tutorial <- ifelse( file.exists(  paste(path,"tutorial/tutorial.Rmd", sep="/") ),
    #3                    btn(path,"blue","tutorial"),
    ##                    "")
    
    
    homework <- ifelse( file.exists(  paste(path,"homework.Rmd", sep="/") ),
                        btn(path,"indigo","homework", as.nb=TRUE),
                        "")
    
    buttons <- c(buttons, 
                 paste(vid,slides,narriative, homework))
    
  }
  
  titles <- nonunderscorification( files )
  
  df <- data.frame( Topic=titles,
                    Connections = buttons
  )
  
  df %>%
    knitr::kable(format="html", 
                 escape=FALSE, 
                 caption = caption) %>%
    kableExtra::kable_styling( bootstrap_options = c("striped", 
                                                     "hover", 
                                                     "condensed", 
                                                     "responsive"),
                               full_width = FALSE,
                               position = "left",
                               fixed_thead = T) -> theTable
  
  return( theTable )
  
}

```


## The R Language

The following components focus on fundamental aspects of the `R` statistical language and should be applicable to anyone who is working with `R` on a regular basis.

```{r}
files <- list.dirs( path="r_language", recursive=FALSE )
makeTable( files, caption="Content lectures for R language materials.")
```

## Markdown  

Markdown is a tremendously powerful tool for bringing your data, your analyses, and the manner in which you communicate them into tigher collaboration.  Learning markdown will make your life *so* much easier, you will wonder how you got along without it up until now... `r emo::ji("smile")`.

```{r}
files <- list.dirs( path="markdown", recursive=FALSE )
makeTable( files, caption="Merge data, communication, and visualization into a single framework using RMarkdown and numerous extensions thereof.")
```


## Visualizing Data

The ability to visualize data is one of the most fundamental components of data literacy.  The following topics 

```{r}
files <- list.dirs( path="visualization", recursive=FALSE )
makeTable( files, caption="Content lectures for visualizing data in R.")
```


## Spatial Data in R


```{r}
files <- list.dirs( path="spatial", recursive=FALSE )
makeTable( files, caption="Content lectures for working with geospatially referenced data (points, lines, polygons, and rasters) in R.")
```





## Data Manipulation

This is where the rubber meets the road in data analysis.  By far, the vast amount of effort we expend in our daily work is getting data into a format that makes sense to us, that can surface various insights, and be inserted into our analysis workflows.  

```{r}
files <- list.dirs( path="manipulation", recursive=FALSE )

makeTable( files, caption="These are methods based upon data manipulation, <i>sensu lato</i>.")
```


## Statistical Models

Once we get the data and configure it properly, we can test specific hypotheses by the application of statistical models.

```{r}
files <- list.dirs( path="models", recursive=FALSE )
makeTable( files, caption="A broad collection of general statiatical models as applied to data in R.")
```




<hr>

This file last updated: `r date()`.