---
title: "Choropleth Maps"
output: html_notebook
---

## Base R Plotting

A choropleth map displays divided geographic regions that are colored in relation to some numerical value.  The `maps` library provides some basic maps for you to do some quick visualization.  Here is a map of the continental US.  

```{r}
library(maps)
map('state', projection = "polyconic" )
```


I grabbed some census data from the US Census site for each state and saved it as a CSV file.  It is currently hosted on a Github site I use for all my ENVS teaching and we can read it directly from Github using the URL.

```{r message=FALSE}
library(readr)
url <- "https://raw.githubusercontent.com/dyerlab/ENVS-Lectures/master/lectures/spatial/basic_maps/states.csv"
sz <- read_csv( url )
head(sz)
```
In this data set, there are 48 states (Hawaii & Alaska are not there) and the District of Columbia.  To color these, I'm going to use a color palette defined in the `RColorBrewer` library that goes from yellow to orange to red.  This is a sequential color ramp (see  `brewer.pal.info` for other palettes).

```{r}
library(RColorBrewer)
colors <- brewer.pal(9,"YlOrRd")
colors
```

Then I'm going stretch these colors out from 9 distinct colors to 49 colors, one for each State and D.C.

```{r}
colors <- colorRampPalette( colors )(49)
colors
```

Then I can `cut` the data into 49 chunks and then assign each state to one of thos



```{r}
groups <- cut( sz$population, 49 )
sz$colors <- colors[ groups ]
head(sz)
```

```{r}
map( 'state', 
     col=sz$colors, 
     fill=TRUE, 
     projection="polyconic")
```