---
title: "Relational Data Homework"
author: YOUR_NAME_HERE
output: html_notebook
---


```{r setup, include=FALSE}
library( tidyverse )
library( knitr )
knitr::opts_chunk$set( warning = FALSE, 
                       error = FALSE,
                       fig.retina = 3)
```

This homework is going to play around with relational joins.  We will use some flight data from 2013 in New York City[^1].

```{r}
library( nycflights13 )
```

This library has five different data frames.  Here is an overview of the raw data.


## The Data 

A table of airline carrier abbreviations (N = `r nrow(airlines)` different carriers).

```{r}
head( airlines )
```

Information on the airports (N = `r nrow(airports)` airports).

```{r}
head( airports )
```
Mostly in the US.

```{r echo=FALSE, fig.cap="Airportts in continental United States.", message=FALSE, warning=FALSE}
if(!require(leaflet)){
  install.packages("leaflet")
}

library( leaflet )
airports %>%
  filter( lon < 0,
          lon > -130,
          lat < 54) %>%
  leaflet( ) %>%
  addMarkers( lat = ~lat,
              lng = ~lon,
              popup = ~faa) %>%
  addProviderTiles( provider="OpenTopoMap")
```

Weather conditions at each airport (a total of `r nrow(weather)` of them!)

```{r}
head(weather)
```
And a list of airplanes (N = `r nrow( planes )`).

```{r}
head( planes )
```

From all of these auxiliary data, we can construct a series of flights which has the following variables in it.

```{r}
names( flights )
```

If you look at the flight table, you will see that several columns of data here are *connected* to the other data frames described above.

`flights` Foreign Key | Data From 
----------------------|----------------
carrier               | airlines
dest                  | airports
origin                | airport
tailnum               | planes 
time_hour             | weather

The reason that these tables have *keys* connecting one to the other is that there is an economy of scale for very large data sets if we partition it into realtional components such as this.


# Questions

Use the tables above to answer the following questions.

1. Which carriers have planes with over 100 flights in 2013 (and how many)?  

```{r}

```




2. During what 48 hour period were the worst delays?  Is there any connection with weather?


```{r}

```






3. What are the five best and five worst airlines for keeping on schedule?

4. Independent of the length of the flight, are you more likely to be delayed by taking a large plane or a small plane? 

5. Do all planes have tail numbers?








[^1]: If you get an error when attempting to load this library in, install it with `install.packages("nycflights13")` and then load it in again. 