---
title: "Relational Data Homework"
author: YOUR_NAME_HERE
output: html_notebook
---


```{r setup, include=FALSE}
library( knitr )
library( chron )
library( lubridate )
library( tidyverse )

knitr::opts_chunk$set( warning = FALSE, 
                       error = FALSE,
                       fig.retina = 3)
```

This homework is going to play around with relational joins.  We will use some flight data from 2013 in New York City[^1].

```{r}
library( nycflights13 )
```

This library has five different data frames.  Here is an overview of the raw data.


## The Data 

A table of airline carrier abbreviations (N = `r nrow(airlines)` different carriers).

```{r}
head( airlines )
```

Information on the airports (N = `r nrow(airports)` airports).

```{r}
head( airports )
```
Mostly in the US.

```{r echo=FALSE, fig.cap="Airportts in continental United States.", message=FALSE, warning=FALSE}
if(!require(leaflet)){
  install.packages("leaflet")
}

library( leaflet )
airports %>%
  filter( lon < 0,
          lon > -130,
          lat < 54) %>%
  leaflet( ) %>%
  addMarkers( lat = ~lat,
              lng = ~lon,
              popup = ~faa) %>%
  addProviderTiles( provider="OpenTopoMap")
```

Weather conditions at each airport (a total of `r nrow(weather)` of them!)

```{r}
head(weather)
```
And a list of airplanes (N = `r nrow( planes )`).

```{r}
head( planes )
```

From all of these auxiliary data, we can construct a series of flights which has the following variables in it.

```{r}
names( flights )
```

If you look at the flight table, you will see that several columns of data here are *connected* to the other data frames described above.

`flights` Foreign Key | Data From 
----------------------|----------------
carrier               | airlines
dest                  | airports
origin                | airport
tailnum               | planes 
time_hour             | weather

The reason that these tables have *keys* connecting one to the other is that there is an economy of scale for very large data sets if we partition it into realtional components such as this.


# Questions

Use the tables above to answer the following questions.

1. Which carriers have planes with over 100 flights in 2013 (and how many)?  

```{r}
flights %>%
  group_by(carrier, tailnum ) %>%
  summarize( flights = length( distance) ) %>%
  filter( flights >= 100 ) %>%
  left_join(airlines, by="carrier") %>%
  group_by( carrier ) %>%
  summarize( Planes = length(tailnum) ) %>%
  left_join( airlines, by="carrier" ) %>%
  arrange( -Planes ) %>%
  select( Carrier = name, `Number of Flights` = Planes) %>%
  knitr::kable( caption="Number of planes by carrier that have over 100 flights in 2013.") %>%
  kableExtra::kable_styling(full_width = FALSE)

  
```




2. During what 48 hour period were the worst delays?  Is there any connection with weather?

```{r}
flights %>% 
  arrange( time_hour ) %>%
  mutate( Julian = yday(time_hour) ) %>%
  select( Julian, arr_delay ) %>%
  group_by( Julian ) %>%
  summarize( Delay = mean( arr_delay, na.rm=TRUE ) ) %>%
  mutate( WindowAve = data.table::frollmean( Delay, n=2 ) ) %>%
  mutate( Maximum = ifelse( WindowAve == max(WindowAve, na.rm=TRUE), TRUE, FALSE)) %>%
  filter( !is.na(Maximum) ) -> delays


most_delay <- month.day.year( delays$Julian[ delays$Maximum == TRUE],
                              origin = c(month=1,day=1,year=2013))


delays %>%
  ggplot( aes(Julian, WindowAve) ) + 
  geom_line() + geom_point( aes(color=Maximum) ) 
```

The 48-hour period with the highest average delay was `r most_delay$month`/`r most_delay$day`.


3. What are the five best and five worst airlines for keeping on schedule?

```{r}
flights %>%
  group_by( carrier ) %>%
  summarize( Delay = mean( arr_delay, na.rm=TRUE) ) %>%
  arrange( -Delay ) %>%
  left_join( airlines, by="carrier") %>%
  select( Carrier = name, Delay) %>%
  knitr::kable( digits=3, caption = "Average delay by carrier in 2013.") %>%
  kableExtra::kable_styling( full_width = FALSE )
```

4. Independent of the length of the flight, are you more likely to be delayed by taking a large plane or a small plane? 

```{r}
flights %>%
  group_by( tailnum ) %>%
  summarize( AveDelay = mean( arr_delay, na.rm=TRUE) ) %>% 
  left_join( planes, by="tailnum" ) %>%
  select( seats, AveDelay ) %>%
  group_by( seats ) %>%
  summarize( Delay = mean( AveDelay) ) %>%
  mutate( Seats = as.numeric( seats ) ) %>%
  ggplot( aes(Seats,Delay) ) + 
  geom_point()  + 
  stat_smooth()
```




5. Do all planes have tail numbers?

```{r}

flights %>%
  filter( is.na(tailnum)) %>%
  select( arr_time ) -> missing_tailnums

any(!is.na(missing_tailnums)  )

```

It looks like all flights with no `tailnum` are canceled flights





[^1]: If you get an error when attempting to load this library in, install it with `install.packages("nycflights13")` and then load it in again. 