---
title: "Data Types Tutorial"
output: 
  learnr::tutorial:
    css: css/tutorial.css
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```


## Background

Welcome to the first interactive tutorial for "Foundations in R", a set of topics that are intended to help everyone get a basic foundation of R understanding prior to jumping into more specific instruction.  This webpage is an interactive tutorial (written in `R` no less) and has both videos and content that will help you gain understanding, experience, and a bit more confidence in using `R` for your anlaytical needs.  You can come back as many times as you like and revisit both the content and self-evaluations presented herein.


**What is R Video Here**


### Variable Assignments

Assignment requires two things (1) you must have a value or object and (2) you must have a variable or object to assign it to.  Instead of using a single equals character for this, `R` uses two characters in a sequence (no spaces between them) to denote *assignment* of a value (or set of values) to an object.  For example, the code below assigns the statement *not a loser* (which is a character data type) to a variable named `dr_dyer` (phew, that was close!).

```{r }
dr_dyer <- "not a loser"
```

Notice how the assignment operator is actually two separate characters, a *less than* sign and a *dash* right next together.  If you read it as a statement, it says, 

> Dr Dyer is assigned the value of 'not a loser'

This approach has *directionality* to it, as it looks a bit like an arrow.  For completeness, the reverse pointing arrow is also permissible.

```{r }
"still not a loser" -> dr_dyer
```

Notice here that we are taking the value and assigning it to the variable.  

<div class="box-red">The only rule here is that the arrow must **always** point towards the entity receiving the data.  </div>

To see the value within a variable, just type the variables name in the R console and hit return. 

```{r }
dr_dyer
```

For conventions, I'll stick mostly with the former throughout all the presentations for simplicity but your own coding style and your desire to craft more readable (and hense less-sucky) code will determine when each is more appropriate.

#### Exercise 

*Here's a simple exercise with an empty code 'chunk' (that is what we call those blocks) provided for you to type some R code.  If you hit the 'Run Code' button, it will evaluate the code and give you an answer.  Write some R code that use the variable `your_name` and contain your name (remember to enclose the text of your name in quotes).  Then on the next line make `R` print out your name (in glorious victory over the computers!!!!!)*

```{r your-name, exercise=TRUE}

```


```{r quiz-assignment, echo=FALSE}
quiz(
  question("Which one of the following will not assign is NOT the value of 23 to the variable `x`?",
           answer("x <- 23", message="The arrow points in the direction of the assignment."),
           answer("x -> 23", message="The arrow points in the direction of the assignment."),
           answer("x <-- 23", correct = TRUE, message="Absolutely.  R will interpret the second dash as a negative sign.  Test for yourself.")
           ),
  caption="Test your knowledge."
)
```



### Equality

Determining if two things are equal is another case where one could use the old `=` character.  R has decided against using it like this for clarity, readability, and general "make-sure-you-do-not-mistype-what-you-want-ability".  Instead of a single equals sign to indicate equality, we have two of them, right next to each other (`==`).


```{r }
2 == 2
dr_dyer == "a loser"
```

In this case, it is true, all values of `2` are identical to `2` and there is no indication that `dr_dyer` is *a loser*.  Notice the responses here are `TRUE` and `FALSE` (both actual logical variables as described below). Also notice how the variables assigned in the chuncks of code above (e.g., the `dr_dyer` part) are available throughout the document!  One of the really cool consequences of using markdown (imho).

For inequality, we do the same thing but instead of two equals symbols, we use an exclamation mark and an equals sign `!=` which is read as *not equal*.

```{r }
2 != 42
```

For more complicated statements, we may need to use the `!` (negation) operator in conjunction with parentheses.  When we wrap some statements within parentheses, the values inside are evaluated *before* the stuff outside (and from left to right).  So, for the expression *is true equal to false*, we get:

```{r}
(TRUE == FALSE)
```

but if we negate that value (e.g., make it the opposite) using the `!` in front of the parentheses

```{r}
!(TRUE == FALSE)
```




#### Exercise 

*Use the chunk of code below to answer the following questions.*

```{r ex-equality, exercise=TRUE}

```


```{r quiz-equality1, echo=FALSE}
quiz(
  question("What is the result of 23 == 23?",
           answer("YES", message="Sorry, Yes is not a result for equality operators in R"),
           answer("FALSE", message="For most cases of 23, they are currently equal to other values of 23."),
           answer("TRUE", correct = TRUE, message="Absolutely. 23 is totally the same as 23"),
           allow_retry = TRUE
           ),
  caption=""
)
```







### Naming Variables

While we are on the subject of typecase, it is important to say a few things about how we pick our variables.  If you want your "future self" to absolutely loath your "present self" then use variables like:

```{r }
x <- 21
x2 <- 42
x43 <- "bob"
```

which tells the person reading the code *absolutely nothing* about what the variable is representing.  

Conversely, if you want to *make friends and influence people* through your code, try naming variables in a way that represents what they are being used for so that when you come back to this code later, it makes sense.

```{r }
the_meaning_of_life <- 42
dr_dyers_loser_status <- FALSE
depthOfTheRiver <- 32.345
number.of.tacos.eaten <- 4
```

Here are the base rules for making a variable name:  

  - It *must* start with a letter.  
  - It cannot have any spaces in it.  
  - It can have numbers or a few symbols within it (only if they actually do some good at making it more readable, like a period or an underscore).
  - They are *all* case sensitive (e.g., `Bob`, `BOb`, `BoB`, `boB`, `bOB`, and `Robert` are all completely different variable names).



### Quiz

Test your knowledge on the topics covered in this section.

```{r quiz-variable-names}
quiz(
  question("Which of the following examples are not valid variable names?",
           random_answer_order = TRUE,
           answer("x"),
           answer("23andMe", correct=TRUE),
           answer("The Dog", correct=TRUE),
           answer("thisIsCamelCase"),
           answer("BoBRobertbOb"),
           answer("X$%B", correct=TRUE)
           ),
  caption = "Evaluate your knowledge."
)
```











## Numeric Data



> Numerical data types contain all numerical represenations.

By far, the most common kind of data we use in our analyses is numerical data. This may represent measured things like `height`, `snout-vent length` (whatever that is), `depth`, `age`, etc.  In data analysis, we commonly take (or obtain) measurements from several items and then try to characterize them using summaries and visualization.

In `R`, the numerical data type can be defined as:

```{r }
X <- 42
```

Notice how the numerical value of `42` is assigned to the variable named `X`.  To have `R` print out the value of a particular variable, you can type its name in the console and it will give it to you.

```{r }
X
```


### Operators

Numeric types have a ton of normal operators that can be used just like a regular calculator.  For the variables `x` and `y`

```{r }
x <- 10
y <- 23
```

we have the *arithmatic operators*:

```{r }
x + y
x - y
x * y
x / y
```

You have the *exponentials*:

```{r }
## x raised to the y
x^y

## the inverse of an exponent is a root, here is the 23rd root of 10
x^(1/y)
```

The *logrithmics*:

```{r }
## the natural log
log(x)

## Base 10 log
log(x,base=10)
```

And the *modulus operator*:

```{r }
y %% x
```

If you didn't know what this one is, don't worry.  The modulus is just the *remainder after division* like you did in grade school.  The above code means that *23 divided by 10 has a remainder of 3*.  I include it here just to highlight the fact that many of the operators that we will be working with in `R` are created by more than just a single symbol residing at the top row of your computer keyboard.  There are just too few symbos on the normal keyboard to represent the breath of operators.  The authors of `R` have decided that using combinations of symbols to handle these and you will get used to them in not time at all.


### Exercise 

*Here is an interactive code chunk.  Use it to answer the following questions*

```{r age-height, exercise=TRUE, lines=5, exercise.eval=FALSE}
age <- 50
height_in_meters <- 2



```


```{r quiz-operators, echo=FALSE}
quiz(
  question("What is the result of `age %% height_in_meters?",
           answer("25"),
           answer("100"),
           answer("0", correct = TRUE)
           ),
  question("What is the results of the variable `age` raise to the power of `height_in_meters`?",
           answer("25"),
           answer("250"),
           answer("2500", correct = TRUE),
           answer("25000")
           ),
  caption = ""
)
```








## Logical Data






### Section 1



#### Exersize 











## Factor Data











