---
title: "R Grammar"
output: 
  html_notebook:
      css: "envs543-styles.css"
---

```{r setup, include=FALSE}
  library(emo)
```

Because `R` is a full language, rather than a `point-and-click` interface, it is important that we understand the general grammar and how to use it.


# Assignments & Interactivity

This document is an [R Notebook](https://dyerlab.github.io/ENVS-Lectures/notebooks/slides.html).  It is written in [R Markdown](https://dyerlab.github.io/ENVS-Lectures/markdown/slides.html) which allows you to mix in regular text (like this) with `R` code, analyses, and output as a single document.  This is a particularly helpful approach to working with data as you keep your code, raw data, analyses, and output all within a single area—thus making your research more [Reproducible](https://dyerlab.github.io/ENVS-Lectures/reproducible_research/slides.html).  If you would like to download a copy of these data, select `Download Rmd` from the `Code` menu at top right of this page and it this web document will produce the raw markdown you can view in [RStudio](https://rstudio.com).

In this section, I take a brief detour to cover some of the most basic aspects of interacting with `R`.  I'll assume that you are interacting with it in the *Console*.  First, the `R` language specification makes a distinction between the assignment of a value to a variable and the query as to the equality of two objects.  In doing so, we do not use a single equals sign (e.g., `=`) for either of these operations.  

## Variable Assignment

Assignment requires two things (1) you must have a value or object and (2) you must have a variable or object to assign it to.  Instead of using a single equals character for this, `R` uses two characters in a sequence (no spaces between them) to denote *assignment* of a value (or set of values) to an object.  For example, the code below assigns the statement *not a loser* (which is a character data type) to a variable named `dr_dyer` (phew, that was close!).

```{r}
dr_dyer <- "not a loser"
```

Notice how the assignment operator is actually two separate characters, a *less than* sign and a *dash* right next together.  If you read it as a statement, it says, 

> Dr Dyer is assigned the value of 'not a loser'

This approach has *directionality* to it, as it looks a bit like an arrow.  For completeness, the reverse pointing arrow is also permissible.

```{r}
"still not a loser" -> dr_dyer
```

Notice here that we are taking the value and assigning it to the variable.  

<div class="box-red">The only rule here is that the arrow must **always** point towards the entity receiving the data.  </div>

For conventions, I'll stick mostly with the former throughout all the presentations for simplicity but your own coding style and your desire to craft more readable (and hense less-sucky) code will determine when each is more appropriate.

## Equality

Determining if two things are equal is another case where one could use the old `=` character.  R has decided against using it like this for clarity, readability, and general "make-sure-you-do-not-mistype-what-you-want-ability".  Instead of a single equals sign, we have two of them, right next to each other.


```{r}
2 == 2
dr_dyer == "a loser"
```

In this case, it is true, all values of `2` are identical to `2` and there is no indication that `dr_dyer` is *a loser*.  Notice the responses here are `TRUE` and `FALSE` (both actual logical variables as described below). Also notice how the variables assigned in the chuncks of code above (e.g., the `dr_dyer` part) are available throughout the document!  One of the really cool consequences of using markdown (imho).

## Case Sensitivity

This is totally going to bite you in the back side as you learn R, because generally humans are sloppy typists.  The `R` language is case-sensitive!  This means that the variables are different if they are not typed **exactly** the same.

```{r}
x <- 21
X <- 42
x == X
```
## Variable Naming
 

While we are on the subject of typecase, it is important to say a few things about how we pick our variables.  If you want your "future self" to absolutely loath your "present self" then use variables like:

```{r}
x <- 21
x2 <- 42
x43 <- "bob"
```

which tells the person reading the code *absolutely nothing* about what the variable is representing.  

Conversely, if you want to *make friends and influence people* through your code, try naming variables in a way that represents what they are being used for so that when you come back to this code later, it makes sense.

```{r}
the_meaning_of_life <- 42
dr_dyers_loser_status <- FALSE
depthOfTheRiver <- 32.345
number.of.tacos.eaten <- 4
```

Here are the base rules for making a variable name:  

  - It *must* start with a letter.  
  - It cannot have any spaces in it.  
  - It can have numbers or a few symbols within it (only if they actually do some good at making it more readable).
  - They are *all* case sensitive (e.g., `Bob`, `BOb`, `BoB`, `boB`, `bOB`, and `Robert` are all completely different variable names).


All variables in `R` have an intrinsic knowledge of what *class* of variable they belong to (e.g., numeric, logical, etc.).  You can ask a variable about it's class using the function `class()`.

```{r}
class( the_meaning_of_life )
class( dr_dyers_loser_status )
```

These will become more important later on when we look at coercion.



