---
title: "Analysis of Mean Values"
output: html_notebook
---

For this homework, we are looking at the data from the Rice Rivers Center.  Here we will be testing individual, pairs, and complete sets of mean values using `t.test` and `aov` values.  Start by loading in the raw data.

```{r}
source("getRiceData.R")
data <- getRiceData()
```


Complete the following tests.

1. Test the null hypothesis that the average overnight air temperatures (say between 7am & 7pm) on all Mondays in January & February were above freezing.  *Describe your results verbally with the aid of the statistics, visualization, or whatever you feel will help you make your point*.

```{r}
data %>%
  filter( month( Date ) < 3 ) %>% 
  filter( Weekday  == "Monday" ) %>%
  filter( hour( Date ) >= 7 & hour( Date ) <= 19 )  %>%
  select( AirTemp ) %>%
  t.test( mu = 0, alternative = "greater" )
```

2. Are weekend water temperatures in February warmer than weekend air temperatures? *Describe your results verbally with the aid of the statistics, visualization, or whatever you feel will help you make your point*.

```{r}
data %>%
  filter( Weekday %in% c("Saturday","Sunday"), 
          month( Date ) == 2 ) -> weekend

data %>%
filter( !(Weekday %in% c("Saturday","Sunday")), 
          month( Date ) == 2 ) -> weekday

t.test( x = weekend$H2O_TempC, y= weekday$AirTemp, aalternative = "less" )
```

3. Are there differences in the average rain recorded each weekeay at the Rice Rivers Center?  *Describe your results verbally with the aid of the statistics, visualization, or whatever you feel will help you make your point*.

```{r}
fit.aov <- aov( Rain ~ Weekday, data=data)
fit.aov
anova( fit.aov )
fit.tuk <- TukeyHSD( fit.aov )
fit.tuk
plot( fit.tuk )
```



