---
title: "Data Containers"
output: 
  html_notebook:
      css: "envs543-styles.css"
---

All of the collection types below can be applied to any of the data types discussed [previously](https://dyerlab.github.io/ENVS-Lectures/data_types/narriative.nb.html).


## Vectors of Values

It is rather rare that we work with a single entity for any data type.   The main syntax in `R` used to make a collection of objects is to enclose them into the `c()` function (*c* for *c*ombine and since we use it so often, making it as short as possible—anything we can do to type less is good for reducing the number of errors in our code).  

Here is how we can get a vector of `numeric` values.

```{r}
x <- c(1,2,3,4,5,6,7)
x
```

The variable `x` contains a number of numeric types and is a single object.

```{r}
length( x )
```



### Sequences

It would really suck if we had to type in all the elements into every array we want.  Say we want a vector of 100 numbers, it would be really likely if we had to type all the numbers `1, 2, 3, ... , 99, 100` (and quite boring `r emo::ji("zzz")`).  The main way we will get data into `R` is by loading it from files or databases, which we will get to later, but for now let's look at how we can specifically create sequences of values with a little bit of code.

The first is the colon operator.  The syntax is `X:Y` which will produce the values from `X` to `Y` (inclusive on both ends).  In `R` we use it like this:

```{r}
y <- 3:8 
y
```

This is just a shortcut for the more general `seq()` function (*seq* is short for *sequence*).  This function takes a value to start at and one to end with (inclusive just like above).

```{r}
y <- seq(3,8)
y
```

These are the *required* arguments.  By default, the `seq()` function counts by $1.0$ (either up or down, depending upon the magnitude of the first and second argument).

```{r}
seq(8,3)
```

If we want it to count by some other interval, we can tell it by passing the *optional argument* `by=`.

```{r}
y <- seq(3,8,by=0.25)
y
```



### Indexing

Even though these objects are solitary, they have several elements.  For example, the object `y` above has `r length(y)` entities in it.  To get to these individual entities within the `y` we use a square bracket notation to indicate that we want to get an item *within* the variable and must indicate which element (or elements) we are interested in getting.  The numeric indices within the brackets are 1-based (e.g., the first element is 1, the second is 2, etc.).  So the code

```{r}
y[4]
```

returns the 4$^{th}$ element in the vector.  This notation can be used to *get* as well as *set* values within the vector.

```{r}
y[2] <- 42
y
```

### Slices

In addition to getting a single entity from a vector, we can also grab a sequence of values by using a set of indices within the square brackets.  For example, if I wanted the first five entries of `y` I could ask for it as:

```{r}
y[1:5]
```









