---
title: "Cartography"
author: "Charis Deadwyler"
date: "2/10/2021"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE
)
library(sf)
library(maps)
library(units)
library(rgdal)
library(rgeos)
library(GGally)
library(raster)
library(mapproj)
library(leaflet)
library(rnaturalearth)
library(rnaturalearthdata)
library( readr )
library( tidyverse )
library( lubridate )
library(knitr)
library(htmltools)

library(scales)
library(tidycensus)
library(tigris)
```


# Cartography

"The art or technique of making maps"

### Factors that effect design
- Objective
- Audience
- Reality and generalization
- Map scale
- Technical limits

### Basic elements of a Map

- Title
- Legend
- Scale
- North arrow
- Additional information


### Effective Maps

- Communicates a clear message
- Appropriate projection
- Data at appropriate level of generalization
- Clear symbology
- Symbols that match the data
- Good figure-ground organization
- Good visual hierarchy


# Common Types of Maps

### Dot-Density
- Spatial distribution and quantity
https://github.com/tarakc02/tarakc02.github.io/blob/master/dot-density/index.md

```{r}

options(tigris_use_cache = TRUE)


v19 <- load_variables(2019, "acs5", cache = TRUE)

v19 %>%
  mutate(table = str_extract(name, "^.+_")) %>%
  filter(str_detect(concept, "EDUCATIONAL ATTAINMENT")) %>%
  select(table, concept) %>%
  distinct() %>%
  print(n = Inf)


#census_api_key("21cac509ab0c7c0ac7b33b020b91a9a4efe40c74", install = TRUE)

acs <- get_acs("tract", table = "B15003", cache_table = TRUE,
               geometry = TRUE, state = "51", county = "760",
               year = 2019, output = "tidy")


acs <- acs %>%
  mutate(
    id = str_extract(variable, "[0-9]{3}$") %>% as.integer
  ) %>%
  filter(id > 1) %>%
  mutate(education = case_when(
    id %>% between(2, 16) ~ "No HS diploma",
    id %>% between(17, 21) ~ "HS, no Bachelors",
    id == 22 ~ "Bachelors",
    id > 22 ~ "Post-Bachelors"
  )) %>%
  group_by(GEOID, education) %>%
  summarise(estimate = sum(estimate))


```

Creating Dots

```{r}
acs_split <- acs %>%
  filter(estimate > 50) %>%
  split(.$education)


generate_samples <- function(data)
  suppressMessages(st_sample(data, size = round(data$estimate / 10)))

points <- map(acs_split, generate_samples)
points <- imap(points,
               ~st_sf(data_frame(education = rep(.y, length(.x))),
                      geometry = .x))
points <- do.call(rbind, points)


```


```{r}

points <- points %>%
  group_by(education) %>%
  summarise()

points <- points %>%
  mutate(education = factor(
    education,
    levels = c("No HS diploma", "HS, no Bachelors",
               "Bachelors", "Post-Bachelors")))

points <- points %>%
  mutate(n_points = map_int(geometry, nrow))
```

Plotting

```{r}

water <- tigris::area_water("51", "760")
roads <- tigris::roads("51", "760")
counties <- tigris::counties(state = "51") 
rva <- counties %>%
  filter(COUNTYFP == 760)
```


```{r, out.width= '100%'}
ggplot() +
  geom_sf(data = rva, fill = "white", lwd = 1) +
  geom_sf(data = water, fill = "lightblue", colour = "lightblue") +
  geom_sf(data = roads, colour = "grey") +
  geom_sf(data = points,
          aes(colour = education,
              fill = education),
          size = .5) +
  scale_color_brewer(type = "div", palette = 4,
                     name = "Education") +
  scale_fill_brewer(type = "div", palette = 4) +
  theme_minimal() +
  ggtitle("Distribution of educational attainment in Richmond, Virginia (2019)",
            "1 dot equals 10 people") +
  guides(fill = "none")

#ggsave("DotDensity.png")
```
### Proportional Symbols
- Size of symbols is related to the value of phenomenon represented

```{r}
v19 %>%
  mutate(table = str_extract(name, "^.+_")) %>%
  filter(str_detect(concept, "TOTAL POPULATION")) %>%
  select(table, concept) %>%
  distinct() %>%
  print(n = Inf)

acs <- get_acs("county", table = "B01003", cache_table = TRUE,
               geometry = TRUE, state = "51",
               year = 2019, output = "tidy")
points <- acs %>%
  st_centroid(of_largest_polygon = TRUE)
```



```{r}


ggplot() +
  geom_sf(data = counties, fill = "lightblue", color = "lightblue") +
  geom_sf(data = acs, fill = "grey95") +
  geom_sf(data = points, aes(size = estimate), color = "lightsalmon2") +
  theme_minimal() +
  ggtitle("Population of Virginia Counties",
          "2019") +
  labs( size = "Population") +
  scale_size_continuous(labels = comma)

#ggsave("ProportionalSymbols.png")
 
```


### Choropleth
- Data is colored or shaded base on an attribute

```{r}

acs <- acs %>%
  mutate(quant = cut(estimate, quantile(estimate, seq(0,1, len = 10)),
                     include.lowest = TRUE))

acs$population <- log(acs$estimate)

ggplot() +
  geom_sf(data = counties, fill = "lightblue", color = "lightblue") +
  geom_sf(data = acs, aes(fill = quant), color = "black", lwd = .5) +
  scale_fill_brewer(type = "seq", palette = 4)

ggplot() +
  geom_sf(data = counties, fill = "lightblue", color = "lightblue") +
  geom_sf(data = acs, aes(fill = population), color = "black", lwd = .5) +
  scale_fill_continuous(type = "viridis", 
                        name = "Population (log transformed)")+
  theme_minimal() +
  ggtitle("Population Density of Virginia Counties",
          "2019")
#ggsave("Choropleth.png")
```


### Isoline
- Lines connect points knows or estimated to have equal vales
- Used in maps of temperature, precipitation, air pressure, climate classifications, ect
- Topographic maps: type of isoline map depicting elevation

```{r}
library(ggisoband)

class(volcano)
names(volcano)

volcano3d <- reshape2::melt(volcano)
names(volcano3d) <- c("x", "y", "z")

ggplot(volcano3d, aes(x,y, z=z)) +
  geom_isobands(aes(fill = stat(zmin)), color = NA) +
  scale_fill_viridis_c() +
  coord_cartesian(expand = FALSE) +
  theme_minimal()

```

```{r}
#DEM <- raster("USGS_one_meter_x28y416_VA_Sandy_2014.tif")

#DEM_5m <- aggregate(DEM, fact = 5)

#DEM_5m <- writeRaster(DEM_5m, filename = file.path("DEM_5m.tif"))

DEM_5m <- raster("DEM_5m.tif")

DEM_pts <- rasterToPoints(DEM_5m, spatial = TRUE)
DEM_df <- data.frame(DEM_pts)
DEM_df$DEM_5m <- ifelse(DEM_df$DEM_5m <0, 0, DEM_df$DEM_5m)


ggplot(DEM_df, aes(x=x, y=y, z=DEM_5m)) +
  geom_isobands(aes(fill = stat(zmin)), color = NA) +
  scale_fill_viridis_c()

#ggsave("Isoline.png")

ggplot(DEM_df, aes(x=x, y=y, z=DEM_5m)) +
  geom_isobands(aes(color = stat(zmin)), fill = NA) +
  scale_color_viridis_c() +
  theme_minimal()

#ggsave("Isoline_lines.png")
```




# Basic Map Making

```{r}
library("ggplot2")
theme_set(theme_bw())
library("sf")

library("rnaturalearth")
library("rnaturalearthdata")

```


note - `returnclass` default is `sp`
```{r}

world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)

```


```{r}
ggplot(world) +
  geom_sf() -> p

p

setwd("C:/Users/Charis/Documents/Spring2021/GISinR/GISinR")

#ggsave("world.png")
```

### Title, subtitle & axis labels

```{r}
setwd("C:/Users/Charis/Documents/Spring2021/GISinR/GISinR")

p +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("World Map",
          subtitle = paste0("(" ,length(unique(world$name)), "countries)" )) -> p

p



ggsave("world_title.png")
```
* `paste()` vs `paste0()`

### Color

```{r}
setwd("C:/Users/Charis/Documents/Spring2021/GISinR/GISinR")

p +
  geom_sf(color = "black", fill = "paleturquoise") -> p

p

ggsave("color.png")
```

`color` -> outline
`fill` -> filled color


```{r}

setwd("C:/Users/Charis/Documents/Spring2021/GISinR/GISinR")

p +
  geom_sf(aes(fill = pop_est)) +
  scale_fill_viridis_c( trans = "sqrt") -> p

p


ggsave("palette.png")
```

`trans = ` transformation of data, in this case square root

`option = ` A-E, default is D

```{r}

setwd("C:/Users/Charis/Documents/Spring2021/GISinR/GISinR")

beetle_url <- "https://raw.githubusercontent.com/dyerlab/ENVS-Lectures/master/data/Araptus_Disperal_Bias.csv"

read_csv(beetle_url) %>%
  st_as_sf( coords=c("Longitude","Latitude"), crs=4326 ) %>%
  st_union() %>%
  st_buffer(dist = 1) %>%
  st_bbox() -> bounds

read_csv(beetle_url) %>%
  st_as_sf(coords=c("Longitude","Latitude"), crs = 4326) -> beetle

ggplot(world) +
  geom_sf() +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("Beetles") +
  coord_sf( xlim = c(-116, -108), ylim = c(22, 31), expand = FALSE) -> p

p

ggsave("arrow.png")

```


Scale bar and North arrow
package `ggspatial`

```{r}
setwd("C:/Users/Charis/Documents/Spring2021/GISinR/GISinR")

library(ggspatial)

p+
  annotation_scale(location = "br", style = "bar") +
  annotation_north_arrow(location = "br", width = unit(.5, "cm"),
                         height = unit(.75, "cm"),
                         pad_y = unit(0.75, "cm"),
                         pad_x = unit(0.5, "cm"),
                         style = north_arrow_orienteering(
                           text_size = 8
                         )) -> p

p

ggsave("arrow2.png")
```



```{r}
ggplot(world) +
  geom_sf() +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("Beetles") +
  coord_sf( xlim = c(-118, -85), 
            ylim = c(10, 35), 
            expand = FALSE) +
  annotation_scale(location = "bl", 
                   style = "bar") +
  annotation_north_arrow(location = "bl", width = unit(.5, "cm"),
                         height = unit(.75, "cm"),
                         pad_y = unit(0.75, "cm"),
                         pad_x = unit(0.5, "cm"),
                         style = north_arrow_orienteering(
                           text_size = 8
                         )) -> p

p
```


changed extent to see more countries


### Adding names
```{r}
library(ggrepel)

world_points<- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))

my_points <- world_points %>%
  filter( X > -118 & X < -85) %>%
  filter( Y > 10 & Y < 35)

nrow(my_points)
summary(world_points$Y)

ggplot() +
  geom_sf(data=world, fill = "seashell") +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("Mexico", subtitle = "& Friends") +
  coord_sf( xlim = c(-118, -85), 
            ylim = c(10, 35), 
            expand = FALSE) +
  annotation_scale(location = "bl", 
                   style = "bar") +
  annotation_north_arrow(location = "bl", width = unit(.5, "cm"),
                         height = unit(.75, "cm"),
                         pad_y = unit(0.75, "cm"),
                         pad_x = unit(0.5, "cm"),
                         style = north_arrow_orienteering(
                           text_size = 8
                         )) +
  geom_text_repel(data= my_points,aes(x=X, y=Y, label=name), 
                  color = "royalblue4", fontface = "bold",
                  size = 3.5,
                  point.padding = 0,
                  box.padding = unit(.5, "cm"))   +
  annotate(geom = "text",
               x = -90, y = 26, 
               label = "Gulf of Mexico", 
               fontface = "italic", 
               color = "grey22", 
               size = 4) +
    annotate(geom = "text",
               x = -113, y = 21, 
               label = "Pacific Ocean", 
               fontface = "italic", 
               color = "grey22", 
               size = 4) +
  theme(panel.background = element_rect(fill = "azure")) -> Mexico_map

#ggsave("Mexico.png")


Mexico_map
```



```{r}
mexico <- ne_countries(scale = "medium", type = "countries", country = "Mexico", returnclass = c("sf"))

ggplot() +
  geom_sf(data = mexico,  aes(fill = "subunit")) +
  geom_sf(data = beetle, color = "darkred") +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("Beetles") +
  coord_sf( xlim = c(-116, -108), ylim = c(22, 31), expand = FALSE) +
  theme(panel.background = element_rect(fill = "azure"))

names(mexico)

```


### Layouts

Side-by-Side  
```{r}
library(ggthemes)

ggplot(data = world) +
  geom_sf(aes(fill = region_wb)) +
  ggtitle("World") +
  labs(fill = "Region", size = 2) +
  geom_rect(xmin = -116, xmax = -108, ymin = 22, ymax = 31, 
     fill = NA, colour = "black", size = 1) +
  scale_fill_viridis_d() +
  theme(panel.background = element_rect(fill = "azure"),
     panel.border = element_rect(fill = NA)) -> world_map

world_map



```

```{r}
ggplot() +
  geom_sf(data=world, fill = "seashell") +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("Mexico") +
  coord_sf( xlim = c(-118, -85), 
            ylim = c(10, 35), 
            expand = FALSE) +
  annotate(geom = "text",
               x = -90, y = 26, 
               label = "Gulf of Mexico", 
               fontface = "italic", 
               color = "grey22", 
               size = 1) +
    annotate(geom = "text",
               x = -113, y = 21, 
               label = "Pacific Ocean", 
               fontface = "italic", 
               color = "grey22", 
               size = 1) +
  theme(panel.background = element_rect(fill = "azure")) -> Mexico
```



```{r}
setwd("C:/Users/Charis/Documents/Spring2021/GISinR/GISinR")

library(cowplot)

plot_grid(world_map, Mexico, nrow = 1, rel_widths = c(2.3, 1)) -> p
p

ggsave("grid_plot2.png")
```

Inset Map

```{r}
VA_map 
ggplot(counties) +
  geom_sf()

ggplot() +
  geom_sf(data = counties, fill = "lightblue", color = "lightblue") +
  geom_sf(data = acs, fill = "grey95") +
  theme_minimal() -> VA_map

ggplot() +
  geom_sf(data = rva, fill = "grey95", lwd = 1) +
  geom_sf(data = water, fill = "lightblue", colour = "lightblue") +
  geom_sf(data = roads, colour = "grey")+
  coord_sf(datum = NA) -> RVA_map

VA_map +
  annotation_custom(
    grob = ggplotGrob(RVA_map), 
      xmin = -83.6,
      xmax = -75 + (-77.4 - (-75))/2.5,
      ymin = -77.3,
      ymax = -37.6 + (-75.1 - 37)/2.5
  )


st_bbox(counties)
st_bbox(rva)

(ratioRVA <- (37.60281 - 37.44655) / (-77.38551 - (-77.60117)))

ggdraw(VA_map) +
  draw_plot(RVA_map,
            width = 0.25, height = 0.25* 10/4 * ratioRVA, 
        x = .15, y = .5)

```

```{r}

arrowA <- data.frame(x1 = 18, x2 = 25, y1 = 3.45, y2 = 3.55)

ggplot() +
  geom_sf(data = rva, fill = "grey95", lwd = 1) +
  geom_sf(data = water, fill = "lightblue", colour = "lightblue") +
  geom_sf(data = roads, colour = "grey")+
  annotate("text", x = -77.575, y = 37.62, label= "Richmond", size = 3)+
  theme_void() +
  theme(panel.grid.major = element_line(colour = gray(0.5), linetype = "dashed", 
         size = 0.5), panel.background = element_rect(fill = "white"), 
         panel.border = element_rect(fill = NA)) +
  coord_sf(datum = NA) -> RVA_map

ggdraw(xlim = c(0, 35), ylim = c(0, 7.5)) +
  draw_plot(VA_map, x = 0, y = 0, width = 25, height = 7.5) +
  draw_plot(RVA_map, x = 24, y = .5, width = 10, height = 7)+
  geom_segment(aes(x = x1, y = y1, xend = x2, yend = y2), data = arrowA, 
        arrow = arrow(), lineend = "round") -> arrow_box


#ggsave("arrow_box.png")


```



