---
title: "Functions"
author: YOUR_NAME_HERE
output: html_notebook
---

> Functions are a way to compartmentalize chuncks of code so you can use them over and over.

For this homework, I would like you to create a function called `get_rice()` and it will do the following:

- Load in the csv file from google docs.  
- Make columns for Date, Month, & Weekday.  
- Convert Temperature & Length units to metric version.  
- Drop all duplicate data.  
- Make column names readable and usable.  
- Return the data frame to the user.  


I'll get you started with the URL and the stub for the function.  If you want to look at the raw data

```{r}
# Put your code here for the function.

get_rice <- function() {
  url <- "https://docs.google.com/spreadsheets/d/1Mk1YGH9LqjF7drJE-td1G_JkdADOU0eMlrP01WFBT8s/pub?gid=0&single=true&output=csv"
  
  # Put all your code here
  
  
}
```



When done, you should be able to call the function as: 

```{r}
rice <- get_rice() 
summary( rice )
```



## Put it into a script file.

Copy the code for the function above and put it into a *.R script file and save it in the root directory of your project for this class.  Now you can use it whenever we need to load in that particular data file.

