---
title: "Regression In-Class Activities"
output: html_notebook
---

```{r setup, include=FALSE}
library( tidyverse )
library( nycflights13 )
```



Since regression analysis is such a fundamental component of inferences, I thought it would be a good idea to try to work through a few more examples to give you some more practice.  

For this we will use some data on flights into and out of Newark (EWR), JFK, and La Guardia airports in NYC metropolitan area.

```{r}
summary( flights) 
```

We also have some data related to local weather conditions.

```{r}
summary( weather )
```

I'm going to make a synthetic data set consisting of flights out of JFK for the *ExpressJet* airlines.  Since the weather data is on the hour, I also make a synthetic varible `Month.Day.Hour` so that I can join the weather data to this one.

```{r}
flights %>%
  filter( origin == "JFK", carrier == "EV") %>%
  mutate( DateTime = paste(month,day,hour, sep=".")) %>%
  select( DateTime, 
          `Departure Delay` = dep_delay,
          `Departure Time` = sched_dep_time) %>%
  droplevels() -> df.jfk 
summary( df.jfk )
```

Now I'll do the same with the weather data.

```{r}
weather %>%
  filter( origin == "JFK") %>% 
  mutate( DateTime = paste(month,day,hour, sep=".")) %>%
  select( -origin, -year, -month, -day, -hour, -time_hour ) -> df.weather 
summary( df.weather )
```

Now, I'm going to `merge` these two `data.frame` objects using the common `key` I made `DateTime`.

```{r}
merge( df.jfk, df.weather, by="DateTime" ) %>%
  select( -DateTime ) -> df 

summary( df )
```


# Activity

What features of this data set may impact the departure delay for flights coming out of JFK on *ExpressJet* Airlines?






